How to write a PhD in Biological Sciences: a guide for the uninitiated

### 2021-07-20 2.05-paragraph.Rmd
Add files via upload
# After copy edits

## Writing a paragraph {#paragraph}

In the following chapters, I will explain how to write a scientific paper in a formulaic style, as well as the advantages that this has over starting from scratch each time. The suggestion is that once, you sketch out your paper in a 'bullet-point' outline. This outline will give you an idea of the subject in each paragraph, as well as how the paragraphs fit in with each other. The other suggestion was to flesh out this outline by assigning relevant citations to help you remember those main ideas.

Writing paragraphs has a skill all of its own, and the aim of this chapter is to go over the basics of how to put a paragraph together. Other people have also written about this, and you can read an example here.

**This paragraph starts with a topic sentence which contains the principle idea that citations are needed for examples.** The following sentences put this idea into the context of your study. This should be based on examples from existing literature with the corresponding in-text citations provided (Smith et al 2017; Jones 2018). You can use examples in your paragraph to make your point. For examples, Smith and Jones (2017) carried out a study that showed that examples without citations have no credibility for readers. And this was backed up by another study by Jones and Smith (2018), which provided further evidence for the need to provide examples. Once you’ve provided your evidence, and introduced any covariates that you need to, like the use of literature for examples, then you are ready to sum up and link. *After reading this paragraph, the reader will know why citations are needed for examples, and they’ll be ready to read more about the topic of writing a paragraph in more detail.*

In the paragraph above, the topic sentence is bold, the supporting sentences are normal font, and the clincher is in italics.

### Topic sentence
A topic sentence allows the reader to understand quickly the idea/topic you are putting forward in the paragraph. It must be in the context in which you are going to develop the same topic. There’s no point in just mentioning a topic in passing or using it in a different way than you will later.

@@

### Supporting sentences
Supporting sentences convey all the relevant information to the reader. They are going to be statements that are well cited, showing readers where the original ideas came from. Be sure to keep these sentences on topic, and regularly refer back to your outline to make sure that you keep to the original objective of the paragraph. These are the meat of the paragraph and it’s really important to get them right.

If you are writing about differences, then state which way the difference is. For example, if bullfrogs are larger than leaf litter frogs, then say this. Telling your audience that something is different than something else only ends up leaving them guessing about the way in which the things are different. There’s no point in drip dripping this information through. Set it down in as little space as possible so that your reader doesn’t get bored.

@@

Your supporting sentences are best built in the formula of an [argument](argument.html).

### The magic of threes
We can learn a lot about writing convincingly in science by borrowing from storytelling, and one of the greatest lessons in storytelling involves the magic of threes [@angler2020telling].

In putting together your paragraph you can use the magic of threes to break your dialogue into three parts: thesis, antithesis and synthesis (a, b and a+b). Think of it this way: there’s a standard explanation of how a phenomenon that you want to investigate interacts with the variable of your choice (thesis). Then there’s some contradictory evidence that suggests that the standard explanation is not necessarily correct (antithesis). Lastly, there is your approach to investigate exactly how the variable and phenomenon interact which should explain (or at least advance our understanding) of how the variable and the phenomenon are related (synthesis). It may be that this construct should be fleshed out into three paragraphs instead of three sentences within a paragraph.

You won’t be able to construct every paragraph with the magic of threes, but once you have learned this rule, you’ll be able to use it to help emphasise important points in your manuscript that you want the reader to remember as they move forwards. What if your list is two instead of three? Well then you’ll lose the magic, but maybe you can think it through it again? If you have two, then you normally have three by means of the combination of the two (a, b and a+b).

@@

### Use an example
Examples are a very powerful way of conveying ideas in a short amount of space. Don’t replace your paragraph with an example, but do use an example if it shows the reader just what you want. You should be able to do this in a sentence (or two), but if you’re tempted to go on, it’s probably not a good example. Because the first paragraph sets out ideas, it's unlikely that an example there will be a good idea.

### Avoid lists
I’m not a big fan of paragraphs which are simply a long list with little or no thought offered. The worst ones are where there are so many citations along the way that it’s really hard to pick out what is a sentence and what is citation. I understand that it’s important to show precedent and that there is merit in showing how widespread an idea is over taxa or in different disciplines. You never find these lists in journals where words and/or citations are limited, which suggests that you can dispense with them.

@@

### The last sentence of the paragraph: the clincher
Once you’ve conveyed all of the information that you planned to impart in your outline, it’s time for the last sentence. This should conclude the evidence that you’ve provided on your topic. Try not to make it lame. For example: “This shows that little work has been done.” One of my pet hates is authors who insist on telling us that “little work has been done”, as if this was a justification for doing anything. Instead, make it a real clincher about why the topic is important, or how and why you will tackle it. Instead (or as well if you can), you may want this sentence to link on to another paragraph (topic), especially if flow is important at that part of your outline (see below). If you have just constructed a paragraph with thesis, antithesis and synthesis, then your clincher might be simply to underline that your approach in the paper is to investigate the synthesis. Either way, make sure that your last sentence is on topic, and one that sticks in the readers’ minds.

### Above all - read it!
Your paragraph is not finished until you've read it. Reading is an essential part to writing that cannot be emphasised enough. Does your paragraph say what you planned in your outline? If your paragraph and any other text doesn't make sense to you, it sure won't impress anyone else. If you can't bear to read it through immediately, then do it after you've written two or three paragraphs. I suggest that you don't wait until you've finished the manuscript. Rather get the text right as you go along.

### How does the paragraph fit into the flow?
So now we’ve gone over the formula, it’s time to take a step back and look again at the paragraph in the context of your outline. Remember that the paragraph represents a single subject, but that it is still just part of the manuscript as a whole and you need that to flow from beginning to end. This means that it’s not enough to write each paragraph in isolation, but to think of the way in which they link together as a whole, especially when you are trying to make a logical argument.

You might get a comment from me about missing logic in the argument or a lack of flow in a paragraph. This usually comes from not having a good outline (see below). You should be using the outline to provide the logic in the flow of your argument, flesh it out with the relevant references. If you already know this well, then writing the paragraph will just be a question of styling the outline. Remember that there is a purpose to your paragraph. You have an argument to make the reader understand. Don't be tempted to include information that is distracting or moving away from the flow of that argument.

@@

### Seeding ideas
The introduction sets out the established literature in order to put your study in context, but your discussion provides you with an opportunity to present new ideas, or to turn and twist existing ideas in a new way. Once you’ve got a good idea what these are, I like to seed the introduction with hints as to what these might be. Sowing seeds early in a manuscript will provide the reader with hints as to where you are going. Writing these seeds as questions is a really good way of sowing them into an introduction. You can then go on to answer them (if only partly) in the discussion. Beware though, there’s no point in asking a major question in the introduction to which your data has no relevance!

### Breaking the rules
Just as in the other instructions on formulaic paper writing, when writing a paragraph you shouldn’t feel totally constrained so that you can’t break the rules. Breaking the rules can set you free, and much of what you read that really stands out will do this. However, it’s much easier to break the rules and get it wrong, than break them and get it right. The idea of this chapter was to help you get started, not to communicate with those who are already writing great stuff. So if you're already great, don't break it by doing any of this!

### Some common paragraph mistakes
- Don't start your paragraph by linking back to the previous point. Start with your new topic sentence. If you aren't ready to change the topic, you aren't ready for a new paragraph.
- Don't start the paragraph with an unsubstantiated or over-inflated claim (keep it calm and within the bounds of what you are citing). Similarly, don't start with a trivial caveat. And don't start with a list or long string of citations.
- Don't let your paragraph get too long. It is said that 250 words is the threshold. If you haven't covered what you wanted to in your outline within 250 words, then you need to rethink your outline for that paragraph! Definitely don't split a paragraph that's too long into two!
- If you started without an outline, it's never too late to go back and write an outline. Simply write down what you are trying to say (as a set of bullet points), and why (see below for an example).
- Your paragraph should have more than one citation in it. If you find yourself citing the same paper several times in the same paragraph, then there's something wrong with your outline, and the way that you are reading and citing papers (see below). Remember that what you write is about making your own argument from the literature, and not copying others arguments.

@@

- You should not start a sentence with an abbreviated genus. Instead write the genus name in full.
- You should not start a sentence with a numeral. If you have to use the number as the first word, write it out. Otherwise, rearrange your sentence. 
- Make sure that acronyms are spelt out on first use (this may be needed for each section)

>>>>

# Writing a paragraph {#paragraph}

Writing paragraphs has a skill all of its own, and the aim of this chapter is to go over the basics of how to put a paragraph together.

**This paragraph starts with a topic sentence which contains the principle idea that citations are needed for examples.** The following sentences put this idea into the context of your study. This should be based on examples from existing literature with the corresponding in-text citations provided (Smith et al. 2017; Jones 2018). You can use examples in your paragraph to make your point. For examples, Smith and Jones (2017) carried out a study that showed that examples without citations have no credibility for readers. And this was backed up by another study by Jones and Smith (2018), which provided further evidence for the need to provide examples. Once you’ve provided your evidence, and introduced any covariates that you need to, like the use of literature for examples, then you are ready to sum up and link. *After reading this paragraph, the reader will know why citations are needed for examples, and they’ll be ready to read more about the topic of writing a paragraph in more detail.*

In the paragraph above, the topic sentence is **bold**, the supporting sentences are normal font, and the clincher is in *italics*.

## Topic sentence
A topic sentence allows the reader to understand quickly the idea/topic you are putting forward in the paragraph. It must be in the context in which you are going to develop the same topic. There’s no point in just mentioning a topic in passing or using it in a different way than you will later.

@@

## Supporting sentences
Supporting sentences convey all the relevant information to the reader. They are going to be statements that are well cited, showing readers where the original ideas came from. Be sure to keep these sentences on topic, and regularly refer back to your outline to make sure that you keep to the original objective of the paragraph. These are the meat of the paragraph and it’s really important to get them right.

If you are writing about differences, then state which way the difference is. For example, if tulips are larger than lillies, then say this. Telling your audience that something is different than something else only ends up leaving them guessing about the directionality. There’s no point in drip dripping this information through. Set it down in as little space as possible so that your reader doesn’t get bored.

@@

Your supporting sentences are best built in the formula of an [argument](#argument).

## The magic of threes
We can learn a lot about writing convincingly in science by borrowing from storytelling, and one of the greatest lessons in storytelling involves the magic of threes [@angler2020telling].

In putting together your paragraph, you can use the magic of threes to break your dialogue into three parts: thesis, antithesis and synthesis (a, b and a+b). Think of it this way: there’s a standard explanation of how a phenomenon that you want to investigate interacts with the variable of your choice (thesis). Then there’s some contradictory evidence that suggests that the standard explanation is not necessarily correct (antithesis). Lastly, there is your approach to investigate exactly how the variable and phenomenon interact which should explain (or at least advance our understanding) of how the variable and the phenomenon are related (synthesis). It may be that this construct should be fleshed out into three paragraphs instead of three sentences within a paragraph.

You won’t be able to construct every paragraph with the magic of threes, but once you have learned this rule, you’ll be able to use it to help emphasise important points in your manuscript that you want the reader to remember as they move forwards. What if your list is two instead of three? Well, then you’ll lose the magic, but maybe you can think it through it again? If you have two, then you normally have three by means of the combination of the two (a, b and a+b).

@@

## Use an example
Examples are a very powerful way of conveying ideas in a short amount of space. Don’t replace your paragraph with an example, but do use an example if it shows the reader just what you want. You should be able to do this in a sentence (or two), but if you’re tempted to go on, it’s probably not a good example. Because the first paragraph sets out ideas, it's unlikely that an example there will be a good idea there.

## Avoid lists
I’m not a big fan of paragraphs which are simply a long list with little or no thought offered. The worst ones are where there are so many citations along the way that it’s really hard to pick out what is a sentence and what is citation. I understand that it’s important to show precedent and that there is merit in showing how widespread an idea is over taxa or in different disciplines. You never find these lists in journals where words and/or citations are limited, which suggests that you can dispense with them.

@@

## The last sentence of the paragraph: the clincher
Once you’ve conveyed all of the information that you planned to impart in your outline, it’s time for the last sentence. This should conclude the evidence that you’ve provided on your topic. Try not to make it lame. For example: “This shows that little work has been done.” One of my pet hates is authors who insist on telling us that “little work has been done”, as if this was a justification for doing anything. Instead, make it a real clincher about why the topic is important, or how and why you will tackle it. Instead (or as well if you can), you may want this sentence to link on to another paragraph (topic), especially if flow is important at that part of your outline (see below). If you have just constructed a paragraph with thesis, antithesis and synthesis, then your clincher might be simply to underline that your approach in the paper is to investigate the synthesis. Either way, make sure that your last sentence is on topic, and one that sticks in the readers’ minds.

## Above all - read it!
Your paragraph is not finished until you've read it. Reading is an essential part of writing that cannot be emphasised enough. Does your paragraph say what you planned in your outline? If your paragraph and any other text doesn't make sense to you, it sure won't impress anyone else. If you can't bear to read it through immediately, then do it after you've written two or three paragraphs. I suggest that you don't wait until you've finished the manuscript. Rather get the text right as you go along.

## How does the paragraph fit into the flow?
So now we’ve gone over the formula, it’s time to take a step back and look again at the paragraph in the context of your outline. Remember that the paragraph represents a single subject but that it is still just part of the manuscript as a whole and you need that to flow from beginning to end. This means that it’s not enough to write each paragraph in isolation, but to think of the way in which they link together as a whole, especially when you are trying to make a logical argument.

You should be using the outline to provide the logic in the flow of your argument, fleshed out with the relevant references. If you already know this well, then writing the paragraph will just be a question of styling the outline. Remember that there is a purpose to your paragraph. You have an argument to make the reader understand. Don't be tempted to include information that is distracting or moving away from the flow of that argument.

@@

## Seeding ideas
The introduction sets out the established literature in order to put your study in context, but your discussion provides you with an opportunity to present new ideas, or to turn and twist existing ideas in a new way. Once you’ve got a good idea of what these are, I like to seed the introduction with hints as to what these might be. Sowing seeds early in a manuscript will provide the reader with hints as to where you are going. Writing these seeds as questions is a really good way of sowing them into an introduction. You can then go on to answer them (if only partly) in the discussion. Beware though, there’s no point in asking a major question in the introduction to which your data has no relevance!

## Breaking the rules
Just as in the other instructions on formulaic paper writing, when writing a paragraph, you shouldn’t feel totally constrained so that you can’t break the rules. Breaking the rules can set you free, and much of what you read that really stands out will do this. However, it’s much easier to break the rules and get it wrong, than break them and get it right. The idea of this chapter was to help you get started, not to communicate with those who are already writing great stuff. So if you're already great, don't break it!

## Some common paragraph mistakes
- Don't start your paragraph by linking back to the previous point. Start with your new topic sentence. If you aren't ready to change the topic, you aren't ready for a new paragraph.
- Don't start the paragraph with an unsubstantiated or over-inflated claim (keep it calm and within the bounds of what you are citing). Similarly, don't start with a trivial caveat. And don't start with a list or a long string of citations.
- Don't let your paragraph get too long. It is said that 250 words are the threshold. If you haven't covered what you wanted to in your outline within 250 words, then you need to rethink your outline for that paragraph! Definitely don't split a paragraph that's too long into two!
- If you started without an outline, it's never too late to go back and write an outline. Simply write down what you are trying to say (as a set of bullet points), and why (see below for an example).
- Your paragraph should have more than one citation in it. If you find yourself citing the same paper several times in the same paragraph, then there's something wrong with your outline, and the way that you are reading and citing papers (see below). Remember that what you write is about making your own argument from the literature, and not copying others' arguments.

@@

- You should not start a sentence with an abbreviated genus. Instead, write the genus name in full.
- You should not start a sentence with a numeral. If you have to use the number as the first word, write it out. Otherwise, rearrange your sentence. .
- Make sure that acronyms are spelt out on first use (this may be needed for each section)

### 2021-07-20 2.06-argument.Rmd
Add files via upload
# updated references with DOI and discussion table
Add files via upload
# After copy edits

Writing is not straightforward. Your objective is to communicate with a reader, someone who you’ve likely never met and will never meet. You need to communicate highly complex information. But more than that you need them to see things as you do. You need to provide them with your reasoning and your argument, and have it make sense to them; preferably to the point where they agree with you.

To communicate, you need to start from common ground. The beginning of your introduction starts with the most general concepts in the context in which you are writing. The context depends on your audience, and this in turn relates to the particular journal that you are writing for. Even if you are writing a thesis, you should pick out a target journal for each of the chapters. Once you’ve established the common ground, you need to carry the reader towards the hypothesis or question that you propose. The easiest way to do this is to make use of a logical argument:

A series of statements that introduce a starting premise, provide evidence for and against that premise (perhaps adding in an example that makes your point), point out what missing information would allow reaching a better understanding of said premise, and logically conclude that what you are doing is going to fill this gap.

@@

The following science argumentation model (Table \@ref(tab:argument)) is modified from Cope et al [-@cope2013science]. Try using this model to guide your writing of a paragraph of your introduction (or discussion) where you attempt a logical argument.

Table: (\#tab:argument) **A framework for a scientific argument.** Modified from Cope et al (2013), this model argument can be followed to provide as a way of getting started.

| *Structure*  | *Explanation*  |
| ------------ | -------------- |
| A position statement / question / hypothesis / theory / problematic | This could be the big idea in your manuscript, or one of several competing concepts that you are introducing. The context in which you are writing might mean that this idea theory needs no introduction (e.g. the theory of evolution in the journal Evolution), but you need to be confident that your audience will understand what you are proposing. Remember to cite the person who came up with the idea. |
| Claim 1 | One potential explanation or interpretation of the original idea. Evidence: Literature that agrees with this interpretation (could include an example) Reasoning: Your justification that links the evidence to the claim or interpretation |
| Claim 2 | Another potential explanation, of the previous interpretation of the original idea | *Evidence*: Literature that agrees with this interpretation (could include an example) *Reasoning*: Your justification that links the evidence to the claim or interpretation |
| Counter - claims | Other possible interpretations or counter claims. |
| Conclusions | Evaluation: your judgement on weighing up the evidence for the idea. |

If your logical argument doesn’t arrive immediately, don’t force it. Give it some thinking time to let it settle in. I find that logical arguments develop best when I’m not sitting at my desk writing. They come when I’m reading other papers (especially those with well written arguments), taking a deep, hot bath, or when I’m out running in the mountains. If you are more social, talk through your logical argument with others, perhaps over a cup of coffee. It really helps to keep the ideas turning in your head. Don’t cut yourself off from work thoughts when you are away from work.

There are other shorter forms that might suit you better. Consider another classic form of the scientific argument: “compare and contrast” which allows you as a writer to quickly familiarise your reader with some key examples. Another sentence structure to consider is explaining cause and effect. For practical wording examplesof both of these classic arguments, [see part 2](phrasebank.html).

## Arguments in Latin
From time to time you may come across comments about arguments that you’ve made in your work made in Latin. Why do people use Latin and not simply say it in English? What do these phrases actually mean? While you may think that these people are being pompous or trying to demonstrate some scholarly superiority by quoting Latin at you (and you might be right), there is some important reasoning to consider both in the meanings of these argumentum (as they are known in Latin), and the fact that they are as old as Roman scholars, and often much older. This then is the point, that some arguments are faulty, and that this has been known for thousands of years.

Here I have provided some of these arguments, and the reasons why they may appear in your thesis in writing biological sciences.

@@

_Non sequitur_ - “It does not follow”: suggests that one statement in your argument is not linked to the next. This is also known as a logical fallacy, a formal fallacy or a deductive fallacy (for a full list of fallacy statements and their explanations, see here).

@@

**Correlation does not equal causation** - That two events or variables are related when in fact they are co-incidental.

**Absence of evidence is not evidence of absence** - The idea that positive evidence of the existence of a relationship is superior to a lack of the same evidence. This is also known as a Type II error (false negative), where the lack of finding a significant relationship does not lead to a conclusion that the relationship does not exist. Later we discuss how a [lack of finding of a significant result](transparency.html) leads to confirmation bias in science.

**Observational selection** -"The root of all superstition is that men observe when a thing hits, but not when it misses." Francis Bacon [-@bacon1620novum]. This refers to the cherry picking of studies that support your hypothesis while omitting those that do not support it. Later we see how this can be abused when writing a review or even the introduction to your chapter.

@@

* Did you understand it?
* Was it written in a conventional style (as in the table (\#tab:argument))
   + If not, how was the style broken with and did this improve or detract from the understanding?

In this writing book, I concentrate on providing formulaic approaches to writing, because these are by far the easiest ways for inexperienced writers. However, I encourage you to learn and experiment with writing styles as you become more experienced. [Critical reading](criticalreading.html) is one of the best ways to learn about alternative writing styles (see here): reading is probably the best way of improving your writing. Alternatively, and especially if the above hasn’t clicked with you, you can read more about writing a scientific argument (see for example [here](https://msu.edu/course/iah/231b/gifford/sciaargs.htm)). And there’s plenty more out there.

>>>>

Writing is not straightforward. Your objective is to communicate with a reader, someone who you’ve likely never met. You need to communicate highly complex information. But more than that you need them to see things as you do. You need to provide them with your reasoning and your argument, and have it make sense to them; preferably to the point where they agree with you.

To communicate, you need to start from common ground. The beginning of your introduction starts with the most general concepts in the context in which you are writing. The context depends on your audience, and this, in turn, relates to the particular journal that you are writing for. Even if you are writing a thesis, you should pick out a target journal for each of the chapters. Once you’ve established the common ground, you need to carry the reader towards the hypothesis or question that you propose. The easiest way to do this is to make use of a logical argument:

A series of statements that introduce a starting premise, provide evidence for and against that premise (perhaps adding in an example that makes your point), point out what missing information would allow reaching a better understanding of said premise, and logically conclude that what you are doing is going to help fill this gap.

@@

The following science argumentation model (Table \@ref(tab:argument)) is modified from Cope et al. [-@cope2013science]. Try using this model to guide your writing of a paragraph of your introduction (or discussion) where you attempt a logical argument. For a more in depth treatment of using arguments in science, see Toulmin [-@toulmin2003uses].

Table: (\#tab:argument) **A framework for a scientific argument.** Modified from Cope et al. (2013), this model argument can be followed to provide as a way of getting started.

| **Structure**  | **Explanation**  |
| ------------ | -------------- |
| A position statement / question / hypothesis / theory / problematic | This could be the big idea in your manuscript, or one of several competing concepts that you are introducing. The context in which you are writing might mean that this idea theory needs no introduction (e.g. the theory of evolution in the journal Evolution), but you need to be confident that your audience will understand what you are proposing. Remember to cite the person who came up with the idea. |
| Claim 1 | One potential explanation or interpretation of the original idea. Evidence: Literature that agrees with this interpretation (could include an example) Reasoning: Your justification that links the evidence to the claim or interpretation. |
| Claim 2 | Another potential explanation, of the previous interpretation of the original idea | *Evidence*: Literature that agrees with this interpretation (could include an example) *Reasoning*: Your justification that links the evidence to the claim or interpretation |
| Counter-claims | Other possible interpretations or counter-claims. |
| Conclusions | Evaluation: your judgement on weighing up the evidence for the idea. |

If your logical argument doesn’t arrive immediately, don’t force it. Give it some thinking time to let it settle in. I find that logical arguments develop best when I’m not sitting at my desk writing. They come when I’m reading other papers (especially those with well-written arguments), taking a deep, hot bath, or when I’m out running in the mountains. If you are more social, talk through your logical argument with others, perhaps over a cup of coffee. It really helps to keep the ideas turning in your head. Don’t cut yourself off from work thoughts when you are away from work.

There are other shorter forms that might suit you better. Consider another classic form of the scientific argument: “compare and contrast” which allows you as a writer to quickly familiarise your reader with some key examples. Another sentence structure to consider is explaining cause and effect. For practical wording examples of both of these classic arguments, [see part 2](phrasebank.html).

## Arguments in Latin
From time to time you may come across comments in Latin about arguments that you’ve made in your work. Why do people use Latin and not simply say it in English? What do these phrases actually mean? While you may think that these people are being pompous or trying to demonstrate some scholarly superiority by quoting Latin at you (and you might be right), there is some important reasoning to consider both in the meanings of these argumentum (as they are known in Latin), and the fact that they are as old as Roman scholars, and often much older. This then is the point, that some arguments are faulty, and that this has been known for thousands of years.

Here I have provided some of these arguments, and the reasons why they may appear in your thesis in biological sciences.

@@

_Non sequitur_ - “It does not follow”: suggests that one statement in your argument is not linked to the next. This is also known as a logical fallacy, a formal fallacy or a deductive fallacy.

@@

**Correlation does not equal causation** - That two events or variables are related when in fact, they are co-incidental.

**Absence of evidence is not evidence of absence** - The idea that positive evidence of the existence of a relationship is superior to a lack of the same evidence. This is also known as a Type II error (false negative), where the lack of finding a significant relationship does not lead to a conclusion that the relationship does not exist. Later we discuss how a [lack of finding of a significant result](transparency.html) leads to [confirmation bias](#TypeI) in science.

**Observational selection** - "The root of all superstition is that men observe when a thing hits, but not when it misses." Francis Bacon [-@bacon1620novum]. This refers to the cherry picking of studies that support your hypothesis while omitting those that do not support it. Later we see how this can be abused when writing a review or even the introduction to your chapter.

@@

* Did you understand it?
* Was it written in a conventional style (as in Table \@ref(tab:argument))
   + If not, how was the style broken with and did this improve or detract from the understanding?

In this writing book, I concentrate on providing formulaic approaches to writing, because these are by far the easiest ways for inexperienced writers. However, I encourage you to learn and experiment with writing styles as you become more experienced. [Critical reading](criticalreading.html) is one of the best ways to learn about alternative writing styles: reading is probably the best way of improving your writing. Alternatively, and especially if the above hasn’t clicked with you, you can read more about writing a scientific argument (see for example [here](https://msu.edu/course/iah/231b/gifford/sciaargs.htm)). And there’s plenty more out there.

### 2021-07-20 2.07-storytelling.Rmd
Add files via upload
# After some small edits over the week
Add files via upload
# After copy edits

## Storytelling in science? {#storytelling}

@@

“No storytelling” is a comment that I sometimes make when reading drafts of manuscripts, chapters, and even when editing for journals. What do I mean by this? Stories are deterministic. That is to say that the storyteller has an end in mind when they start telling the story, and the telling is a way to get to their goal. A story that’s ‘pointless’ will frustrate the audience and won’t engender them to listen to that storyteller again. In a good story, reaching that goal will often result in lots of twists and turns with the goal shrouded in mystery until it is revealed. In a teaching story (like a parable), the goal may be overt, such that the audience relates to the narrative and buys into the same conclusion.

@@

### The danger of a teleological argument

Teleological arguments are those which have their end point as the aim. While it is generally acknowledged that this is a good way to tell a story, it can be a trap when it comes to conducting scientific inquiry. Here, we need to approach science in a very different way to storytelling. When I was doing my PhD I was very frustrated as I had the impression that my advisor knew what result he wanted and designed the study to show it. What probably happened is that he was displaying “motivated reasoning”, but the result is known as “confirmation bias”. When an experiment failed to meet the expected result, he declared that it had failed. As Frances Bacon so elegantly put it:

>__“Once a man’s understanding has settled on something (either because it is an accepted belief or because it pleases him), it draws everything else also to support and agree with it. And if it encounters a larger number of more powerful countervailing examples, it either fails to notice them, or disregards them, or makes fine distinctions to dismiss and reject them, and all this with much dangerous prejudice, to preserve the authority of its first conceptions.”__ 
>Francis Bacon [-@bacon1620novum]

@@

### SCience is the very opposite of storytelling. Or is it?
I often tend to think of the answer to the question as the goal of the study. That I don’t know what the answer is, doesn’t spoil the story for me. The important thing is asking the question. An unexpected answer might send us back to thinking more about the system that we are studying and result in a greater revelation.

A good example of this is the study published by Becker et al [-@becker2018extreme]. Francois found a (spookily) strong relationship between probability that Rose’s dwarf toadlets survive and local rainfall in the breeding season. More rainfall equals greater survival is what one would automatically think (because frogs like rain), but that wasn’t the result obtained. Francois found that survival increased with less rainfall. It wasn’t until we made the connection between the fact that increased rainfall during the breeding season meant that the toadlets spent more time in puddles, that we understood why their survival was decreasing. As the toadlets aren’t feeding during this time they lose weight, and also expose themselves to increased predation pressure. In a dry year, the toadlets will head back to their subterranean refuges much earlier and continue to pursue ‘safer’ feeding and hiding habits. While we might intuitively feel that a ‘safe toadlet’ is a better life-strategy, reduced rainfall means reduced reproduction, and so results in a failure for toads that don’t manage to pass on their genes that year. The result is a variable life-history with the weather, something that was previously unknown. To me, that’s a great story!

But the scientific paper that was written about this ‘story’ [@becker2018extreme] doesn’t have a narrative style, and doesn’t fit the description of a narrative that we discussed above. Instead, it follows the formula that we set out below: Introduction, Materials and Methods, Results and Discussion. This style does not treat the experiment as a complication on the way to the story’s goal. The structure introduces the rationale behind conducting the experiment, then objectively explains the findings, and lastly discusses their meaning in relation to what is already known. From this study, I can spin the story (above), because I know what the endpoint is. But this storyis just one interpretation of the correlation that was found. [Correlation is not causation](argument.html), and more studies would be needed to test the hypothesis concerning the underlying reasons suggested for the relationship observed. The key differences in the storytelling style from the scientific formula are the absence of a known goal at the start of the process (see a more detailed discussion of this by Yarden Katz, [-@katz2013storytelling] here).

There is a role for stories when communicating science as this increases interest, facilitates understanding, and enhances memory. This is particularly true when communicating science to the public and making it more accessible, but it also applies to interactions between scientists, for example at conferences. The [presentations](#proposal-presentation.html) that tell a story and entertain are those we tend to remember. Not easy, but if we do want to communicate well with each other, then we need to learn the art of storytelling, without compromising our scientific objectivity. In addition, there are lots of lessons from storytelling that we can use within scientific papers, and to which I subscribe wholeheartedly. For example, don’t be boring, don’t [bamboozle](style.html), write an active narrative (not in the passive tense), use the magic of threes, and many more.

@@

### Can storytelling help memory of science?
Some argue that it can, and should be an integral part of scientific writing. In her [blog post](https://www.annaclemens.com/blog/elements-of-a-story), Anna Clemens argues that stories should be used within the scientific context. I'll let you read this for yourselves, and ask you to make up your own minds. If you like this way of thinking, you can get deeper into the depths of story thinking with Angler [-@angler2020telling].

>>>>

# Storytelling in science? {#storytelling}

@@

'No storytelling' is a comment that I sometimes make when reading drafts of manuscripts, chapters, and even when editing for journals. What do I mean by this? Stories are deterministic. That is to say that the storyteller has an end in mind when they start telling the story (i.e. teleological; see Figure \@ref(fig:Storytelling)), and the telling is a way to get to their goal. A story that’s ‘pointless’ will frustrate the audience and won’t engender them to listen to that storyteller again. In a good story, reaching that goal will often result in lots of twists and turns with the goal shrouded in mystery until it is revealed. In a teaching story (like a parable), the goal may be overt, such that the audience relates to the narrative and buys into the same conclusion.

@@

## The danger of a teleological argument

Teleological arguments are those which have their end point as the aim. While it is generally acknowledged that this is a good way to tell a story, it can be a trap when it comes to conducting scientific inquiry. Here, we need to approach science in a very different way to storytelling. When I was doing my PhD I was very frustrated as I had the impression that my advisor knew what result he wanted and designed the study to show it. What probably happened is that he was displaying 'motivated reasoning', but the result is known as '[confirmation bias](#TypeI)'. When an experiment failed to meet the expected result, he declared that it had failed. As Frances Bacon so elegantly put it:

>__“Once a man’s understanding has settled on something (either because it is an accepted belief or because it pleases him), it draws everything else also to support and agree with it. And if it encounters a larger number of more powerful countervailing examples, it either fails to notice them, or disregards them, or makes fine distinctions to dismiss and reject them, and all this with much dangerous prejudice, to preserve the authority of its first conceptions.”__ 
>
>Francis Bacon [-@bacon1620novum]

@@

## Science is the very opposite of storytelling. Or is it?
I often tend to think of the answer to the right question as the goal of the study. That I don’t know what the answer is, doesn’t spoil the story for me. The important thing is asking the question. An unexpected answer might send us back to thinking more about the system that we are studying and result in a greater revelation.

A good example of this is the study published by Becker et al. [-@becker2018extreme]. Francois found a (spookily) strong relationship between probability that Rose’s dwarf toadlets survive and local rainfall in the breeding season. More rainfall equals greater survival is what one would automatically think (because frogs like rain), but that wasn’t the result obtained. Francois found that survival increased with less rainfall. It wasn’t until we made the connection between the fact that increased rainfall during the breeding season meant that the toadlets spent more time in puddles, that we understood why their survival was decreasing. As the toadlets aren’t feeding during this time they lose mass, and also expose themselves to increased predation. In a dry year, the toadlets will head back to their subterranean refuges much earlier and continue to pursue ‘safer’ feeding and hiding habits. While we might intuitively feel that a ‘safe toadlet’ is a better life strategy, reduced rainfall means reduced reproduction, and so results in a 'failure' for toads that don’t manage to pass on their genes that year. The result is a variable life-history with the weather, something that was previously unknown. To me, that’s a great story!

But the scientific paper that was written about this ‘story’ [@becker2018extreme] doesn’t have a narrative style, and doesn’t fit the description of a narrative that is discussed above. Instead, it follows the formula that is set out below: **Introduction, Materials and Methods, Results and Discussion**. This style does not treat the experiment as a complication on the way to the story’s goal. The structure introduces the rationale behind conducting the experiment, then objectively explains the findings, and lastly discusses their meaning in relation to what is already known. From this study, I can spin the story (above), because I know what the endpoint is. But this storyis just one interpretation of the correlation that was found. [Correlation is not causation](argument.html), and more studies would be needed to test the hypothesis concerning the underlying reasons suggested for the relationship observed. The key differences in the storytelling style from the scientific formula are the absence of a known goal at the start of the process (see a more detailed discussion of this by Yarden Katz, [-@katz2013storytelling] here).

There is a role for stories when communicating science as this increases interest, facilitates understanding, and enhances memory. This is particularly true when communicating science to the public and making it more accessible, but it also applies to interactions between scientists, for example, at conferences. The [presentations](#proposal-presentation.html) that tell a story and entertain are those we tend to remember. Not easy, but if we do want to communicate well with each other, then we need to learn the art of storytelling, without compromising our scientific objectivity. In addition, there are lots of lessons from storytelling that we can use within scientific papers, and to which I subscribe wholeheartedly. For example, don’t be boring, don’t [bamboozle](style.html), write an active narrative (not in the passive tense), use the magic of threes, and many more.

@@

## Can storytelling help memory of science?
Some argue that it can, and should be an integral part of scientific writing. In her [blog post](https://www.annaclemens.com/blog/elements-of-a-story), Anna Clemens argues that stories should be used within the scientific context. I'll let you read this for yourselves, and ask you to make up your own minds. If you like this way of thinking, you can get deeper into the depths of story thinking with Angler [-@angler2020telling].
